---
title: "Statistical Modeling with R and Stan"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Schedule

| Session | Time                           | Topics                                                                                                     |
|:-------:|--------------------------------|------------------------------------------------------------------------------------------------------------|
| 1       | Day 1, 09:00-11:00             | **R Math & Programming Refresher**                                                                         |
|         |                                | 1. Good coding practices                                                                                   |
|         |                                | 2. Object types and conversions; slicing and indexing                                                      |
|         |                                | 3. Control structures                                                                                      |
|         |                                | 4. Probability distributions                                                                               |
|         |                                | 5. Linear algebra                                                                                          |
| 2       | Day 1, 11:00-13:00             | **Generalized Linear Models**                                                                              |
|         |                                | 1. GLM basics: Systematic component, link function, family/likelihood; model matrix, coefficient vector    |
|         |                                | 2. The simulation approach                                                                                 |
|         |                                | 3. Quantities of interest (definition, calculation, simulation)                                            |
| 3       | Day 1, 14:00-16:00             | **Bayesian Fundamentals**                                                                                  |
|         |                                | 1. Fundamental concepts: Prior distribution, likelihood, posterior distribution                            |
|         |                                | 2. MCMC Algorithms                                                                                         |
| 4       | Day 2, 09:00-11:00             | **Applied Bayesian Statistics Using Stan: Basics** (Application: Linear Model)                             |
|         |                                | 1. Stan: Language, documentation, and core program blocks                                                  |
|         |                                | 2. The Bayesian workflow                                                                                   |
| 5       | Day 2, 11:00-13:00             | **Applied Bayesian Statistics Using Stan: Extensions** (Application: Binary Choice Models)                 |
|         |                                | 1. Optional Stan program blocks: Functions, transformed data, transformed parameters, generated quantities |
|         |                                | 2. Efficiency tuning                                                                                       |
|         |                                | 3. Processing posterior draws in Stan and R                                                                |
| 6       | Day 2, 14:00-16:00             | **Applied Bayesian Statistics Using Stan: Advanced Modeling**                                              |
|         |                                | 1. Measurement models                                                                                      |
|         |                                | 2. Hierarchical models                                                                                     |
|         |                                | 3. Systems of equations                                                                                    |

## Topic 1

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

