---
title: "Lab: Generalized Linear Models"
---

```{r setup, include=FALSE}
## ---- CRAN Packages ----
## Save package names as a vector of strings
pkgs <-  c("foreign", "MASS", "ggplot2")

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages,
       repos='http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)

## ---- Global Objects ----
gles <- 
  read.dta("https://github.com/denis-cohen/statmodeling/raw/main/data/gles.dta")
```


## Quantities of interest with simulated uncertainty

In this lab session, you will get to apply the concepts from the lecture to a logistic regression of individual vote choices for the AfD in 2017. 

The data set `gles` is once again preloaded; the outcome variable is named `vote_afd`.


## Exercise 1

Consider the following logistic regression:

```{r task1, echo = TRUE}
## Estimate
mod <- glm(vote_afd ~ la_self + east + fem + age,
            data = gles,
            family = binomial(link = "logit"))
```

Your task will be to compute the *average marginal effect* of a one-unit increase in `la_self` (anti-immigration preferences) on the probability of voting AfD.

### Simulate the model coefficients

To simulate the model coefficients, take $S$ draws from $\text{MVN}(\hat\beta, \hat \Sigma)$. $S$ should be a large number, at least 1000. Check the dimensions of the resulting object `b_sim`.

*Hints:* 

- `MASS:mvrnorm()` allows you take random draws from a multivariate normal distribution.
- `coef()` allows you to extract the coefficient estimates.
- `vcov()` allows you to extract the estimate for the variance-covariance matrix of the coefficients.

```{r task1-sim, exercise = TRUE}
S <- 1000L
b_sim <- ...
```

```{r task1-sim-solution}
S <- 1000L
b_sim <- MASS::mvrnorm(S, coef(mod), vcov(mod))
dim(b_sim)
```

### Define your covariate scenarios

We are interested in the *average* marginal effect. 

You thus need two data matrices:

1. `X0`: A model matrix with the observed covariate values of all respondents.
1. `X1`: Same as above, but each respondent's value of `la_self` is increased by 1.

Define the two matrices.

```{r task1-x, exercise = TRUE}

```

```{r task1-x-solution}
## Observed model matrices
X1 <- X0 <- model.matrix(~ la_self + east + fem + age, data = gles)

# Adjust X1: One unit increase in la_self
X1[, "la_self"] <- X1[, "la_self"] + 1
```

### Get your unit-specific marginal effects

1. Calculate vectors of linear predictors for all respondents under both scenarios. *Note:* 
      - Your matrix $\beta^\text{sim}$ is of dimensions $S \times K$. 
      - Your model matrices will e of dimensions $N \times K$. 
      - Transpose and multiply the matrices such that your resulting object should be of dimensions $N \times S$.
2. Apply the inverse logit link function (`plogis()`) to your linear predictors. This transforms the linear predictions to expected values (predicted probabilities).
3. Take the difference between the two objects created in step (2) to get your simulated unit-specific marginal effects.

```{r task1-ume, exercise = TRUE}

```

```{r task1-ume-solution}
# Get vectors for the linear predictor under both scenarios
eta1 <- X1 %*% t(b_sim)
eta0 <- X0 %*% t(b_sim)

# Get expected values (map onto probability scale)
ev1 <- plogis(eta1)
ev0 <- plogis(eta0)

# Get unit-specific marginal effect
me <- ev1 - ev0
```

### Get and summarize your average marginal effects

1. Take the mean (average) across all observations to aggregate your unit-level marginal effects into average marginal effects (*hint*: `apply()`). The resulting object should be a length-$S$ vector that gives you your simulated sampling distribution of the average marginal effect.
2. Summarize the simulated sampling distribution
    - Plot the distribution in a histogram.
    - Calculate its mean or median (a numerical approximation of the point estimate).
    - Calculate its standard deviation (a numerical approximation of the standard error).
    - Constructed a simulation-based 95% confidence interval by taking the 2.5% and 97.5% percentiles of the simulated distribution.
3. Concisely interpret the estimate and its uncertainty.

```{r task1-ame, exercise = TRUE}

```

```{r task1-ame-solution}
# Get average marginal effect
ame <- apply(me, 2, mean)

## Summarize
hist(ame, breaks = 50)
mean(ame)
median(ame)
sd(ame)
quantile(ame, c(.025, .975))
```


## Exercise 2

Based on the same model, we now want to visualize the expected values (predicted probabilities) of voting for the AfD along the scale of anti-immigration preferences while keeping all other variables fixed at their sample means or sample proportions.

For this exercise, we can use the same simulated parameters as before.


### Covariate scenarios and quantities of interest

As in Lab 1, we want to iteratively update our covariate scenario along a fine-grained sequence of values for `la_self` in a loop and store our quantity of interest at the current value in a pre-assigned container.

To do this, follow the following steps:

1. Define an initial vector $\mathbf{x}$ that holds the sample means of all covariates.
1. Define a value sequence `la_vals` of length $L=101$ that ranges from 0 to 10.
1. Define a container matrix of dimensions $L \times S$.
1. In a loop from $1$ to $L$, do the following at each iteration:
    - Overwrite the value of `la_self` in the covariate scenario with the current value of `la_vals`
    - Calculate the simulations of the expected value using the current covariate scenario.

```{r task2-x, exercise = TRUE}

```

```{r task2-x-solution}
# Covariate vector (note: we can use X0 for this task)
x <- apply(X0, 2, mean)

# Value sequence
L <- 101L
la_vals <- seq(0, 10, length.out = L)

# Container
ev_sim <- matrix(NA, L, S)

# Loop
for (i in seq_along(la_vals)) {
  # Overwrite element of x
  x["la_self"] <- la_vals[i]
  
  # Store simulations
  ev_sim[i, ] <- plogis(b_sim %*% x)
}
```


### Summarize the simulated sampling distribution and visualize your finding!

1. To present your finding, first summarize the simulated sampling distribution: At each value of `la_vals`, calculate the median as well as the 2.5% and 97.5% percentiles.
2. Plot the prediction ($y$-axis) with its simulated 95% confidence intervals against the values of `la_vals` ($x$-axis). You can use base plots or `ggplot2`.


```{r task2-viz, exercise = TRUE}

```

```{r task2-viz-solution}
# Summarize
ev_sum <- apply(ev_sim, 1, quantile, c(.5, .025, .975))

# Value sequence
L <- 101L
la_vals <- seq(0, 10, length.out = L)

# Plot
## Auxiliary objects
x_lines <-  seq(0, 10, 2)
y_lines <- seq(0, .35, .05)

## Canvas
plot(
  1,
  1,
  type = 'n',
  main = paste("Probability of voting AfD", 
               "as a function of anti-immigration preferences",
               sep = "\n"),
  axes = F,
  xlab = "Anti-immigration preferences",
  ylab = "Pr(Vote AfD)",
  ylim = range(y_lines),
  xlim = range(x_lines)
)
axis(2, outer = FALSE)
axis(1)
rect(
  min(x_lines),
  min(y_lines),
  max(x_lines),
  max(y_lines),
  col = "gray95",
  border = NA
)
for (y in y_lines)
  segments(min(x_lines),
           y,
           max(x_lines),
           y,
           col = "white",
           lwd = 2)
for (x in x_lines)
  segments(x,
           min(y_lines),
           x,
           max(y_lines),
           col = "white",
           lwd = 2)

## Prediction
polygon(
  c(la_vals, rev(la_vals)),
  c(
    ev_sum[2, ],
    rev(ev_sum[3, ])
  ),
  col = adjustcolor("gray30", alpha.f = .2),
  border = NA
)
lines(la_vals,
      ev_sum[1, ],
      lty = 1,
      col = "gray10",
      lwd = 2)
lines(la_vals,
      ev_sum[2, ],
      lty = 1,
      col = "gray30")
lines(la_vals,
      ev_sum[3, ],
      lty = 1,
      col = "gray30")
```
