---
title: "Lab: Applied Bayesian Statistics Using Stan: Basics"
---

```{r setup, include=FALSE}
## ---- CRAN Packages ----
## Save package names as a vector of strings
pkgs <-  c("rstan", "rstantools", "coda", "dplyr", "foreign")

## Install uninstalled packages
lapply(pkgs[!(pkgs %in% installed.packages())], 
       install.packages,
       repos='http://cran.us.r-project.org')

## Load all packages to library and adjust options
lapply(pkgs, library, character.only = TRUE)

## ---- GitHub Packages ----
devtools::install_github("denis-cohen/regplane3D")
library(regplane3D)

## ---- Global learnr Objects ----
gles <- 
  read.dta("https://github.com/denis-cohen/statmodeling/raw/main/data/gles.dta")

lm_code <- 
"data {
  int<lower=1> N; // num. observations
  int<lower=1> K; // num. predictors
  matrix[N, K] x; // design matrix
  vector[N] y;    // outcome vector
}

parameters {
  vector[K] beta;      // coef vector
  real<lower=0> sigma; // scale parameter
}

transformed parameters {
  vector[N] mu;  // declare lin. pred.
  mu = x * beta; // assign lin. pred.
}

model {
  // priors
  target += normal_lpdf(beta | 0, 10);  // priors for beta
  target += cauchy_lpdf(sigma | 0, 5);  // prior for sigma
  
  // log-likelihood
  target += normal_lpdf(y | mu, sigma); // likelihood
}"

## ---- rstan Options ----
rstan_options(auto_write = TRUE)             # avoid recompilation of models
options(mc.cores = parallel::detectCores())  # parallelize across all CPUs
```


## Exercise

### Context

We now return to our example from Labs 1 and 3: A regression of AfD support on pro-redistribution and anti-immigration preferences.

Unlike Lab 3, we now want to use our posterior draws to generate substantively meaningful quantities of interest: A Bayesian version of the 3D visualization from Lab 1.


### Data

The object `gles` and the Stan program `lm_mod` are pre-loaded in the exercise chunk below.

Generate all required data inputs and collect them in an appropriately named list that matches the declared names in your Stan `data` block.

Please make sure to include `se_self` and `la_self` and the corresponding product term and at least one additional covariate.

```{r data, exercise=TRUE}
# Define data
...

# Data as list
standat <- list(...)
```

### Compilation

Now, compile your Stan program from the character string `lm_code`.

```{r compilation, exercise=TRUE}
lm_mod <- ...
```

### Inference

Estimate your model and store the `stanfit` object

```{r estimation, exercise=TRUE}
lm_est <- ...
```

### Model summary

Print your model summary and see if `Rhat` shows any signs of non-convergence.

```{r summary, exercise=TRUE}

```

### Extract your post-warm-up samples

Now, things are about to get interesting.

Extract the posterior draws of your model parameters. Store the draws for `beta`
in a separate object. Check the dimensions of that object. What do the rows and
columns contain?

```{r extraction, exercise=TRUE}
beta_draws <- ...
```

### Calculate the full posterior distribution of your quantity of interest

We have 1000 posterior draws of each model coefficient (taken from two chains à
1000 post-warm-up draws thinned by a factor of 2).

Now, we want to use these to obtain the full posterior distribution of our
quantity of interest.

For reference, here is how we calculated our quantity of interest in Lab 1:

```{r qoi-lab1, echo = TRUE, eval = FALSE}
# Define value sequences
se_vals <- seq(0, 10, .25)
la_vals <- seq(0, 10, .25)

# Initialize container
pred <- array(NA, dim = c(length(se_vals), length(la_vals), 3L))

# Nested loop for point estimates and confidence bounds
for (i in seq_along(se_vals)) {
  for (j in seq_along(la_vals)) {
    x_star <- c(1, se_vals[i], la_vals[j], se_vals[i] * la_vals[j])
    std_err <- sqrt(as.vector(t(x_star) %*% Sigma %*% x_star))
    pred[i, j, 1] <- x_star %*% b
    pred[i, j, 2] <- pred[i, j, 1] + qnorm(.025) * std_err
    pred[i, j, 3] <- pred[i, j, 1] + qnorm(.975) * std_err
  }
}
```

Three things will be different this time around:

1. We do not want point estimates and confidence bounds, but the full posterior. Adjust the dimensions of the container `pred` accordingly.
1. Previously, all elements of `x_star` (except the leading 1) changed at each iteration of the loop. Now, we must hold our background covariates constant at some value, e.g., at their sample mean. Make sure to change the correct elements of `x_star` at each iteration while holding all others constant.
1. Previously `x_star %*% b` was a simple inner product of two $K$-length vectors. Now we must multiply an $S \times K$ matrix ($S$ draws for $K$ coefficients) with a $K$-length vector $x^{\ast}$ (remember that the arguments must be conformable).

Let's give it a try:

```{r qoi-posterior, exercise=TRUE}
...
```

### Summarize the posterior distribution of your QOI

The object `pred` now holds $S$ posterior draws for the expected value of AfD
Support at each combination of `se_vals` and `la_vals`.

Let's summarize the posterior distribution. For each expected values, take the
50%, 2.5% and 97.5% percentiles.
This is a quantile-based summary of the posterior distribution that gives us
the posterior median and the 95% credible interval.

*Hint:* You can use `apply()` on an array, too! Just set the `MARGIN` argument
to those dimensions you would like to retain. But beware: The dimensions of
the resulting object may be different from what you would intuitively expect.

```{r qoi-summary, exercise=TRUE}
pred_summary <- ...
```

### Visualization

You guessed it: We can now use `regplane3D::plane3D()` on `pred_summary` to
visualize our expected values, much like we did yesterday in Lab 1.

```{r viz, exercise=TRUE, fig.align='center', fig.width = 9, fig.height=6}
# Change array dimensions to make them compatible with plane3D()
pred_summary <- aperm(pred_summary, perm = c(2, 3, 1))

regplane3D::plane3D(
  z = pred_summary,
  x = se_vals,
  y = la_vals,
  main = "Expected AfD Support \n by Redistribution and Immigration Preferences",
  zlab = "Support AfD",
  xlab = "Pro-Redistribution",
  ylab = "Anti-Immigration",
  xnlines = 6L,
  ynlines = 6L
)
```
